---
title: "slcr: R Interface to Altair SLC"
format: gfm
---

## Overview

The `slcr` package provides an R interface to Altair SLC, enabling users to execute SLC code and transfer data between R and SLC environments. This document outlines the installation, configuration, and usage of the package.

## Installation

You can install the development version of slcr from GitHub:

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("username/slcr")
```

## Prerequisites

- R (>= 4.0.0)
- Altair SLC (>= 2026)
- Python module `slc` from Altair SLC installation

## Configuration

Before using the package, you need to configure the Python path where Altair SLC is installed. Let's demonstrate the configuration options:

```{r}
#| eval: false
# Method 1: Before loading the package
options(slc.pythonpath = "/path/to/altair/slc/python")
library(slcr)

# Method 2: In your .Rprofile
# Add this line to your .Rprofile file:
options(slc.pythonpath = "/path/to/altair/slc/python")
```

The default path is `/opt/altair/slc/2026/python` if no custom path is specified.

## Basic Usage

Let's walk through some basic examples of using the package:

```{r}
#| eval: false
# Load the package
library(slcr)

# Initialize a connection
conn <- slc_init()

# Execute SLC code
slc_submit("proc print data=sashelp.class;", conn)
```

### Data Transfer Examples

Here's how to transfer data between R and SLC:

```{r}
#| eval: false
# Create sample data
df <- data.frame(
  x = 1:10,
  y = letters[1:10],
  z = rnorm(10)
)

# Write R data frame to SLC
write_slc_data(df, "mydata", conn)

# Read SLC dataset back into R
result <- read_slc_data("mydata", conn)

# View the results
head(result)
```

## Core Functions

The package provides four main functions:

1. `slc_init()`: Initialize SLC connection
   - Returns a connection object used by other functions
   - Automatically configures Python environment

2. `slc_submit()`: Execute SLC code
   - Accepts SLC code as a character string
   - Returns execution results

3. `write_slc_data()`: Write R data frames to SLC
   - Converts R data frames to SLC datasets
   - Handles data type conversion automatically

4. `read_slc_data()`: Read SLC datasets into R
   - Converts SLC datasets to R data frames
   - Preserves data types where possible

## Error Handling

The package includes robust error handling. Here's an example:

```{r}
#| eval: false
# Example with error handling
tryCatch(
  {
    conn <- slc_init()
    slc_submit("proc print data=nonexistent;", conn)
  },
  error = function(e) {
    message("An error occurred: ", e$message)
  }
)
```

## Contributing

Contributions to the package are welcome! Here's how you can contribute:

1. Report issues on GitHub
2. Submit pull requests
3. Suggest new features
4. Improve documentation

## License

This package is released under the MIT License.

## Session Info

```{r}
#| eval: true
sessionInfo()
```