---
title: "Working with R and SLC in Quarto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with R and SLC in Quarto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE # Don't evaluate any code chunks by default
)
```

## Introduction

The `slcR` package enables seamless integration between R and SLC (SAS Language Compiler) within Quarto documents. This vignette demonstrates how to combine R data manipulation with SLC statistical procedures in a single document.

## Installing the SLC Quarto Extension

The `slcR` package includes a Quarto extension that provides native support for SLC code blocks in Quarto documents. To use SLC code blocks with syntax highlighting and proper rendering, you need to install this extension.

### Installation

First, install the extension in your Quarto project directory:

````{verbatim}
```{r install-extension}
library(slcR)

# Install the SLC Quarto extension in the current directory
install_slc_extension()

# Or install in a specific project directory
# install_slc_extension("path/to/your/quarto/project")
```
````

This will create an `_extensions/slc/` directory in your project with the necessary extension files.

### Quarto Document Setup

Once the extension is installed, add it to your Quarto document's YAML header:

````{verbatim}
```
---
title: "My SLC Analysis"
format: html
filters:
  - slcr
---
```
````

### Alternative: Using the Extension Without Installation

If you prefer not to install the extension locally, you can still use SLC code blocks by registering the `slc` engine directly in your document:

````{verbatim}
```{r setup-engine}
library(slcR)

# Register the SLC engine for this document
knitr::knit_engines$set(slc = slcrR:slc_engine)
```
````

## Getting Started

First, load the slcR package and initialize a connection:

````{verbatim}
```{r init}
library(slcR)

# Initialize SLC connection
conn <- Slc$new()
```
````

## Basic Workflow: R to SLC to R

### Step 1: Prepare Data in R

Start by creating or loading data in R:

````{verbatim}
```{r create-data}
# Create sample data
sales_data <- data.frame(
  region = c("North", "South", "East", "West", "North", "South"),
  quarter = c("Q1", "Q1", "Q1", "Q1", "Q2", "Q2"),
  sales = c(1200, 1500, 1100, 1300, 1400, 1600),
  costs = c(800, 900, 750, 850, 900, 950)
)

# View the data
head(sales_data)
```
````

### Step 2: Send Data to SLC

Use the `input_data` chunk option to make R data available in SLC:

````{verbatim}
```{slc input_data=sales_data}
/* View the data in SLC */
proc print data=sales_data;
  title "Sales Data Overview";
run;
```
````

**Expected Output:**
```
                            The SLC System
                          Sales Data Overview

Obs    region    quarter    sales    costs

 1     North       Q1       1200      800
 2     South       Q1       1500      900
 3     East        Q1       1100      750
 4     West        Q1       1300      850
 5     North       Q2       1400      900
 6     South       Q2       1600      950
```

### Step 3: Perform Analysis in SLC

Create summary statistics and new datasets:

````{verbatim}
```{slc input_data=sales_data, output_data="summary_stats"}
/* Calculate profit and summary statistics */
data sales_with_profit;
  set sales_data;
  profit = sales - costs;
  profit_margin = (profit / sales) * 100;
run;

/* Create summary by region */
proc means data=sales_with_profit noprint;
  class region;
  var sales costs profit profit_margin;
  output out=summary_stats 
    mean=avg_sales avg_costs avg_profit avg_margin
    sum=total_sales total_costs total_profit;
run;
```
````

**Expected Output:**
```
NOTE: There were 6 observations read from the data set WORK.SALES_DATA.
NOTE: The data set WORK.SALES_WITH_PROFIT has 6 observations and 6 variables.
NOTE: There were 6 observations read from the data set WORK.SALES_WITH_PROFIT.
NOTE: The data set WORK.SUMMARY_STATS has 5 observations and 11 variables.
```

### Step 4: Analyze Results in R

The `summary_stats` dataset is now available in R:

````{verbatim}
```{r analyze-results}
# Examine the summary statistics
str(summary_stats)
head(summary_stats)

# Create visualizations
library(ggplot2)

# Filter out the overall summary (_TYPE_=0)
regional_summary <- summary_stats[summary_stats$`_TYPE_` == 1, ]

# Plot average profit by region
ggplot(regional_summary, aes(x = region, y = avg_profit)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Profit by Region",
    x = "Region",
    y = "Average Profit ($)"
  ) +
  theme_minimal()
```
````

## Advanced Workflows

### Multiple Data Exchanges

You can pass data back and forth multiple times:

````{verbatim}
```{r customer-data}
# Start with R data processing
customer_data <- data.frame(
  customer_id = 1:100,
  age = sample(18:80, 100, replace = TRUE),
  income = rnorm(100, 50000, 15000)
)

# Add customer segments
customer_data$age_group <- cut(customer_data$age, 
                              breaks = c(0, 30, 50, 70, 100),
                              labels = c("Young", "Middle", "Senior", "Elder"))
```
````

````{verbatim}
```{slc input_data=customer_data, output_data="customer_analysis"}
/* Perform detailed customer analysis */
proc means data=customer_data;
  class age_group;
  var income;
  output out=customer_analysis
    mean=avg_income
    std=std_income
    min=min_income
    max=max_income;
run;

/* Create income quintiles */
proc rank data=customer_data out=customer_ranked groups=5;
  var income;
  ranks income_quintile;
run;
```
````

````{verbatim}
```{r customer-analysis}
# Continue analysis in R
library(dplyr)

# Merge the analysis results
customer_final <- customer_data %>%
  left_join(customer_analysis, by = "age_group") %>%
  mutate(
    income_z_score = (income - avg_income) / std_income,
    high_value = income > quantile(income, 0.8)
  )

# Summary table
customer_final %>%
  group_by(age_group) %>%
  summarise(
    count = n(),
    avg_income = mean(income),
    high_value_pct = mean(high_value) * 100,
    .groups = "drop"
  )
```
````

### Working with External Data Files

You can also work with external data files:

````{verbatim}
```{r external-data}
# Read data from CSV
if (file.exists("data/survey_data.csv")) {
  survey_data <- read.csv("data/survey_data.csv")
} else {
  # Create sample data for demonstration
  survey_data <- data.frame(
    respondent_id = 1:500,
    satisfaction = sample(1:5, 500, replace = TRUE),
    department = sample(c("Sales", "Marketing", "IT", "HR"), 500, replace = TRUE),
    tenure = sample(1:20, 500, replace = TRUE)
  )
}
```
````

````{verbatim}
```{slc input_data=survey_data}
/* Advanced statistical analysis */
proc freq data=survey_data;
  tables department*satisfaction / chisq;
  title "Satisfaction by Department";
run;

proc corr data=survey_data;
  var satisfaction tenure;
  title "Correlation: Satisfaction vs Tenure";
run;

/* ANOVA */
proc anova data=survey_data;
  class department;
  model satisfaction = department;
  means department / tukey;
  title "ANOVA: Satisfaction by Department";
run;
```
````

## Best Practices

### 1. Data Preparation

Always prepare and validate your data in R before sending to SLC:

````{verbatim}
```{r data-prep}
# Clean and validate data
clean_data <- raw_data %>%
  filter(!is.na(key_variable)) %>%
  mutate(
    key_variable = as.numeric(key_variable),
    category = factor(category)
  ) %>%
  arrange(id)

# Check data quality
summary(clean_data)
```
````

### 2. Error Handling

Use R's error handling when working with SLC results:

````{verbatim}
```{r error-handling}
# Safely read SLC output
tryCatch({
  analysis_results <- conn$get_log()
  
  if (nrow(analysis_results) == 0) {
    warning("No results returned from SLC analysis")
  }
  
}, error = function(e) {
  message("Error reading SLC results: ", e$message)
  analysis_results <- NULL
})
```
````

### 3. Documentation

Document your workflow clearly:

````{verbatim}
```{slc input_data=model_data, output_data="model_results"}
/* 
Purpose: Fit logistic regression model for customer churn prediction
Input: model_data (customer features and churn indicator)
Output: model_results (parameter estimates and fit statistics)
*/

proc logistic data=model_data;
  model churn = age income satisfaction tenure / selection=stepwise;
  output out=model_results p=predicted_prob;
run;
```
````

### 4. ODS Graphics — Route Figures Through `&slcr_gpath`

When producing plots with `PROC SGPLOT`, `PROC SGPANEL`, or any ODS graphics procedure, route output through the `&slcr_gpath` macro variable. slcR creates this per-chunk temp directory automatically and scans it after each chunk — figures written there are captured and embedded without accumulating files in your working directory:

````{verbatim}
```{slc fig-cap="Sales by Region"}
ods html body='' gpath="&slcr_gpath" style=htmlblue;
proc sgplot data=work.sales_data;
  vbar region / response=sales stat=mean;
  yaxis label="Average Sales ($)";
  title "Average Sales by Region";
run;
ods html close;
```
````

Without `gpath="&slcr_gpath"`, SLC writes figures to the document's working directory under generic names (`SGPLOT.png`, `SGPLOT1.png`, etc.). These files persist between renders and can cause stale images to appear on re-render. The `&slcr_gpath` pattern avoids this entirely.

You can add `fig-cap` as either a single string (shared by all figures in the chunk) or a YAML list (one caption per figure):

````{verbatim}
```{slc fig-cap:
  - "Sway speed by visit"
  - "Ellipse area by visit"}
ods html body='' gpath="&slcr_gpath" style=htmlblue;
proc sgplot data=work.sway;
  vbox speed / category=visit;
run;
proc sgplot data=work.sway;
  vbox area / category=visit;
run;
ods html close;
```
````

### 5. Reproducibility

Set up your environment for reproducibility:

````{verbatim}
```{r reproducibility}
# Set seed for reproducible results
set.seed(123)

# Document package versions
sessionInfo()

# Save workspace for later use
save.image("analysis_workspace.RData")
```
````

## Troubleshooting

### Common Issues

1. **Data type mismatches**: Ensure R data types are compatible with SLC
2. **Missing values**: Handle NAs appropriately before sending to SLC
3. **Variable names**: Use valid SLC variable names (no spaces, special characters)

### Debugging Tips

````{verbatim}
```{r debugging}
# Check data before sending to SLC
str(my_data)
summary(my_data)

# Verify SLC connection
if (exists("conn")) {
  message("SLC connection active")
} else {
  conn <- Slc$new()
}

# Check SLC logs for errors
logs <- conn$get_log()
if (length(logs) > 0) {
  cat("SLC Output:\n", paste(logs, collapse = "\n"))
}
```
````

## Conclusion

The `slcR` package provides a powerful bridge between R's data manipulation capabilities and SLC's statistical procedures. By combining both tools in Quarto documents, you can create comprehensive, reproducible analyses that leverage the strengths of both environments.

Key benefits:

- **Seamless data exchange** between R and SLC
- **Reproducible workflows** in a single document
- **Rich visualizations** combining SLC analysis with R graphics
- **Flexible analysis pipelines** using the best tool for each task